---
title: "Supplementary results and figures"
author: 
- name: Mengbo Li
  affiliation: Bioinformatics Division, WEHI
output:
  BiocStyle::html_document:
    toc_float: true
package: BiocStyle
---


```{r setup, include = FALSE}
options(width = 100, digits = 3)
knitr::opts_chunk$set(collapse = TRUE, 
                      comment = "#>", 
                      echo = TRUE, 
                      cache = FALSE, 
                      prompt = FALSE,
                      tidy = TRUE,
                      comment = NA, 
                      message = FALSE, 
                      warning = FALSE, 
                      tidy = TRUE, 
                      tidy.opts = list(width.cutoff = 60),
                      fig.width = 8, 
                      fig.height = 5, 
                      dev = "png")
```

# Overview


# Load packages

```{r}
library(tidyverse)
library(proteoDP)
```

Consider natural cubic splines generated with 1, 3 and 5 degrees of freedom:

```{r}
dfList <- seq(1, 5, 2)
lineColours <- RColorBrewer::brewer.pal(4, "Dark2")
```


# Dataset S1: Sydney heart bank data

This dataset is acquired by DIA and processed with Spectronaut with a spectral library generated from pooled DDA samples. 

## Data summary

```{r}
data("heart")
dim(heart)
heart[1:5, 1:5]
```

The overall proportion of missing data 

```{r}
sum(is.na(heart)) / length(heart)
```

## Empirical logistic splines for detected proportions

```{r}
heartRes <- gatherResults(heart)
```

```{r}
for (i in 1:length(dfList)) {
  if (i == 1)
    plotEmpSplines(heartRes$nuis, X = heartRes$splineFits_params0[[i]]$X, 
                   heartRes$splineFits[[i]]$params, 
                   lineCol = lineColours[i], 
                   jitter.amount = 1/ncol(heart)/2, point.cex = 0.1, 
                   capped = FALSE)
  if (i > 1)
    plotEmpSplines(heartRes$nuis, X = heartRes$splineFits_params0[[i]]$X, 
                   heartRes$splineFits[[i]]$params, 
                   plot.dotts = FALSE, 
                   lineCol = lineColours[i], capped = FALSE)
}
legend("bottomright", legend = paste("df = ", dfList), col = lineColours, lwd = 2, lty = 1)
```

## Reduced deviance compared to an intercept model

```{r}
heartRes$devPlot + 
  ylim(0.95, 1)
```


## Empirical logit-linear curve with capped probabilities

```{r}
heartRes$cappedLinearFit$params
plotEmpSplines(heartRes$nuis, X = heartRes$splineFits_params0[[1]]$X, 
               heartRes$cappedLinearFit$params, 
               lineCol = lineColours[1], 
               jitter.amount = 1/ncol(heart)/2, point.cex = 0.1)
```

## Detection probability curve assuming normal observed intensities

```{r}
heartRes$dpcFit$beta
plotDPC(heartRes$nuis, heartRes$dpcFit, jitter.amount = 1/ncol(heart)/2, point.cex = 0.1)
```






# References

- Li, M., Parker, B.L., Pearson, E., Hunter, B., Cao, J., Koay, Y.C., Guneratne, O., James, D.E., Yang, J., Lal, S. and O’Sullivan, J.F., 2020. Core functional nodes and sex-specific pathways in human ischaemic and dilated cardiomyopathy. *Nature communications*, 11(1), pp.1-12.



# Session information

```{r}
sessionInfo()
```

